### 2025-05-15 docs/article/Sections/state-of-the-art.tex
Chapter 2 - Describe the 3 different idioms in of internal DSL: function sequence, nested and chain. Describe type Safety.

@@

>>>>

DSLs for HTML provide an API where methods or functions correspond to the names of
available HTML elements. 
These methods, also known as \textit{builders}, can be combined in a chain of
calls to mimic the construction of an HTML document in a fluent manner.
Martin Fowler\cite{dslbook} identifies three different patterns for combining
functions to create a DSL: 1) \textit{function sequence}; 2) \textit{nested
function}, and 3) \textit{method chaining}, which are illustrated in the snippets of
Figure~\ref{fig:dsl-idioms}.
\begin{figure}[h]
\centering

% Subfloat 1
\begin{minipage}[c]{0.33\linewidth}
  \centering
  \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html();
 head();
  title();text("JT");end();
 end();
 body();
  p();
   text("Hi JATL");
  end();
 end();
end();
  \end{lstlisting}
  \caption*{(a) Function sequence}
\end{minipage}
\hfil
% Subfloat 2
\begin{minipage}[c]{0.23\linewidth}
  \centering
  \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html(
 head(
  title("ST")
 ),
 body(
  p("Hi ScalaTags")
 )
);
  \end{lstlisting}
  \caption*{(b) Nested function}
\end{minipage}
\hfil
% Subfloat 3
\begin{minipage}[c]{0.32\linewidth}
  \centering
  \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html()
 .head()
  .title().text("HF").__()
 .__()
 .body()
  .p()
   .text("Hi HtmlFlow")
  .__()
 .__()
.__();
  \end{lstlisting}
  \caption*{(c) Method chaining}
\end{minipage}

\caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
\label{fig:dsl-idioms}
\end{figure}

From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the \textit{nested
function} idiom used by \textbf{ScalaTags}, as shown in the snippet of
Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring fewer statements than
the other two cases.
This approach of combining function calls is also utilized by \textbf{Hiccup}
Clojure Library and \textbf{j2html} Java library.
One reason for verbosity in \textit{function sequence} and \textit{method
chaining} approaches is the requirement of a dedicated HTML builder to emit the
closing tag, exemplified by \texttt{end()} in \textbf{JATL} (Figure~\ref{fig:dsl-idioms}.a) and
\texttt{\_\_()} in \textbf{HtmlFlow} (Figure~\ref{fig:dsl-idioms}.c).
But the \textit{nested function} approach comes with a notable drawback: \textbf{they
do not support PSSR} because the sequence of nested functions is evaluated
backward to the order in which they are written. 
In other words, arguments are evaluated before the functions are invoked. 
Taking the example in Figure~\ref{fig:dsl-idioms}.b), the \texttt{title()} function is
first evaluated, and its resulting paragraph becomes the argument for the
\texttt{head()} call, which, in turn, becomes the argument for \texttt{html()},
and so on.
If HTML is emitted as functions are called, it will print tags in reverse order. 
The aforementioned DSLs must collect resulting nodes into an internal data
structure, which is later traversed to produce the HTML output. Therefore, they
cannot progressively emit the output as builders are called.

Two other JVM libraries, \textbf{Groovy MarkupBuilder} and
\textbf{KotlinX.html}, also adopt a \textit{nested function} approach, but they
address the backward evaluation issue by implementing \textit{lazy evaluation}
of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
function literals).
Due to the concise form of expressing lambdas with brackets (i.e.,
\texttt{\{\}}) in both Groovy and Kotlin, translating the template shown in
Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html only
requires replacing parentheses with brackets for parent elements.
\textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API 
as an alternative to its original Java API.

@@

Another key characteristic of such DSLs is \textbf{HTML safety}, which refers to
whether they produce only valid HTML conforming to a well-formed document.
To ensure HTML safety, a DSL API should only allow combining calls to builders
that result in valid HTML. 
However, most HTML DSLs using the \textit{function sequence} or \textit{nested
function} approach cannot ensure HTML safety at compile time.

The \textit{function sequence} approach, as illustrated in the snippet of
Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence of
statements, making it challenging to restrict the order of statements. 
In the \textit{nested function} approach shown in
Figure~\ref{fig:dsl-idioms}.b), variable-length arguments are often used to
allow an undefined number of child elements, which cannot be strongly typed in
every programming language.

KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
evaluation, mitigates this issue through function types with a receiver. 
In this approach, the receiver (i.e., \texttt{this} within the lambda) is
strongly typed and provides a set of methods corresponding to legal child
elements.
This enables KotlinX.html to enforce HTML safety by restricting the available
methods during compile time.

To achieve this KotlinX.html provides HTML builders using \textit{function
literals with receiver}~\cite{kotlinlang}.
In Kotlin, a block of code enclosed in curly braces \texttt{\{...\}} is known as
a \emph{lambda}, and can be used as an argument to
a function that expects a \emph{function literal}.
When we write, for example, \texttt{body \{ div \{ hr() \} \}}, we are
invoking the \texttt{body} function with a lambda as its argument. This
lambda, in turn, calls the \texttt{div} function with another lambda as an
argument that creates a horizontal row (i.e. \texttt{hr}).
Each call to an HTML builder (e.g., \texttt{body}, \texttt{div}, \texttt{hr})
creates the child element within the element generated by the outer function
call.

HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the \texttt{KotlinX.html} 
API, and another that employs the \textit{method chaining} idiom, as illustrated 
in the snippet of Figure~\ref{fig:dsl-idioms}.c).
In this approach, the receiver object is implicitly passed as an argument to
each method call, enabling subsequent methods to be invoked on the result of the
preceding one.
This facilitates the composition of methods, with each call building upon the
other.
Similar to KotlinX.html, HtmlFlow ensures HTML safety, restricting the available
HTML builders and attributes after the dot (\texttt{.}) operator.
